Natural variations in water discharge quantity has an inconsistent impact on sediment transport capacity in subglacial channels

### 2023-10-02 Delaneyetal2023b.tex
integrating appendix and updating text

%% To submit your paper:
\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage{placeins}
\usepackage{amsmath}

@@

\title{Water discharge quantity has a variable impact on sediment transport capacity in subglacial channels}

\authors{Ian Delaney\affil{1},  Andrew J. Tedstone\affil{2},   Mauro A. Werder\affil{3,4},  Daniel Farinotti\affil{3,4}, }

@@

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days to months, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated with increased or decreased water velocity $v$ \cite{alley1997} because the size of the channel $S$ changes on a much longer timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
It follows that an increase in water discharge $Q$ will cause the shear stress across the channel $\tau_t$ to respond with water velocity $v$, as opposed to subaerial channels where channel width $w_c$ evolves as well.

  \begin{figure}[h]
  \centering
    \includegraphics[width=0.8\linewidth]{Fig1.pdf}
    \caption{Sketch for the different responses that subglacial and subaerial channels have to an increase in water discharge.  Water velocity magnitudes in the subglacial, $v_g$, and subaerial, $v_f$, channels are shown by arrow length. $S_g$ and $S_f$ represent the wetted area in subglacial and subaerial channels, respectively. The subglacial channel widths $w_{cg}$ remain unchanged, while the subaerial channel widths $w_{cf}$ evolve. Shear stress $\tau$ is responsible for the mobilization of sediment and increases with the number of makers at the channel bed.
      Note that in the subaerial channel parameterization, channels have rectangular shapes with their width much larger than their depth (Section~\ref{sect:fluv}).}
    \label{fig:cartoon}
  \end{figure}
% \ian{Come back to Daniel's comments}

As a result, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences have been implicitly included in a wide range of available models that  quantify sediment transport from catchments in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>{lane2016,perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channels \cite{alley1997}.
%Furthermore, thorough observations of water velocity and channel width in subglacial channels are limited and, to these authors' knowledge, not contrasted with subaerial conditions at the glacier margins. %\mauro{self-cite here?}

@@

The Leverett Glacier in Greenland (\icesheet) serves as the ice sheet setting.
Water discharge and suspended sediment concentration data were collected roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used from \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.
Two gaps in suspended sediment data of roughly one day in length are interpolated across with a linear spline.

@@

With knowledge of wetted area $S_g$, the shear stress, $\tau_g$, between the water and the channel bed is determined through the Darcy-Weisbach formulation
\begin{equation}
  \label{eq:tau}
  \tau_g=\frac{1}{8}\,f_r\,\rho_w\,v_g^2,
\end{equation}
%
where $v_g = \frac{Q}{S_g}$ is the water velocity.

@@

where $w_{cg}$ is the flat channel floor width:
\begin{equation}
  \label{eq:dh2wc}
  w_{cg} = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S_g}{\beta -\sin \beta}}.
\end{equation}

@@

To parameterize the shear stress of water flowing across sediments in the subaerial channel,  the hydraulics parameterization presented in \citeA{tucker1997} is implemented.
Using mass conservation and the Darcy-Weisbach relationship as well as assuming that the channel is wide so that the hydraulic radius is well approximated by the flow depth, then
the shear stress $\tau_f$ at the river bed is
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q}{w_{cf}} \Big)^{\frac{2}{3}} \,\nabla z_c^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\nabla z_c$ is the channel slope, and $f_p$ is the friction factor for subaerial channels.
Channel width $w_{cf}$ is
\begin{equation}
  \label{eq:wcf}
  w_{cf} = k \, Q^\frac{1}{2},
\end{equation}
%
where $k$ is a constant and $\frac{1}{2}$ is a commonly chosen exponent \cite{leopold1953}.
%\mauro{This is for so called ``regime channels''}
Water velocity, $v_f$, is given by rearranging Equation \ref{eq:tau} as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_p\,\rho_w}}.
\end{equation}
%
As above, the width-integrated shear stress is
\begin{equation}
    \label{eq:tautf}
    \tau_{tf}=w_{cf}\,\tau_f.
  \end{equation}

@@

The parameterizations above are applied to hydrological records from the Fieschergletscher (\alpine) and the Leverett Glacier (\icesheet).
The subglacial and subaerial parameterizations represent the subglacial flow of water exiting a glacier and transitioning to subaerial flow, as it moves through the catchment.
Outputs of the parameterizations are meant to represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these cases, \alpine{}  is exemplified by relatively thin ice thickness ($h_{ice}$: $225$\,\unit{m}), low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by thick ice  ($h_{ice}$: $700$\,\unit{m}), high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).

For each glacier, the parameterizations are first applied to a reference test case for each glacier that assumes  subglacial channel with $\beta=\frac{\pi}{2}$. Friction factors $f_r$ and $f_p$ are tuned so that the parameterizations reproduce reasonable water velocities for both \alpine and \icesheet{}\cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
In both subaerial channels, the bed slope is $\nabla z_c = 0.05$.
Water velocity, shear stress, and width-integrated shear stress in the subglacial and subaerial channels from this formulation are presented below.

Second, to characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels further, $100$  parameterizations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$).
Runs with parameter combinations are culled if their mean subglacial water velocity over the season lies outside of the range between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies outside of the range $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
The standard deviation of an output over the course of the season is used to quantify its variability.

Lastly,  outputs are compared with the sediment discharge records of the glaciers using Spearman rank correlation, that is the correlation with respect to the ordering of values from the outputs and the observations.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.
Outputs and observations, which have a native time resolution of $5$\,\unit{min}, are aggregated across timescales from $15$\,\unit{min} to $15$\,\unit{days} by averaging values one half of the timescale before and after a given point time.

@@

\subsection{Role of water discharge in sediment transport capacity}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig2.pdf}
  \caption{Parameterization outputs resulting from the hydrographs shown in panels a and e for the cases called \alpine (a-d) and \icesheet{} (e-h). Gray (black) lines in a and e represent the subglacial channel size (water discharge).  In other panels, blue lines represent outputs from the subglacial channel, while orange lines represent the subaerial channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows denote some events where the variables peaks in the subglacial channel prior to the subaerial one.
    Insets in e-h show the  peak melt event \icesheet{} denoted by the shaded area.
    Note that the x and y axes are different for \alpine{} and \icesheet{}.
      }
  \label{fig:model_outs}
\end{figure}
Following the parameterizations defined in Sections~\ref{sect:sub_mode}~and~\ref{sect:fluv}, the water velocity, shear stress, and width-integrated shear stress exhibit different seasonal evolutions and peaks. Water velocity and shear stress represent the potential for sediment mobilization, and width-integrated shear stress represents a proxy for the total sediment transport capacity across the channel bed (Figure~\ref{fig:model_outs}).

Over the course of the season, large peaks in water velocity and shear stress, along with width-integrated shear stress, occur in response to increases in water discharge in both \alpine{} and \icesheet{}.
This occurs as the subglacial outputs are generally larger than the subaerial values, thus more efficient at transporting sediment.
In the subglacial channel, peaks generally occur over the period when water discharge increases at the fastest rate (Figure~\ref{fig:model_outs}).
Alternatively, in the subaerial channel, peaks in these sediment transport variables occur when water discharge is highest.
The timing of the greatest rate of change in water discharge occurs before the peaks in water discharge.
Here, sediment could be deposited in the proglacial area as the water depressurizes, then remobilized when sediment transport capacity peaks in the subaerial system.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig3.png}
  \caption{Relationship between water discharge and velocity, shear stress, and width-integrated shear stress for \alpine{} (a-c) and \icesheet (d-f).
    These outputs' relationship with measured sediment discharge is given for \alpine in (g-i) and \icesheet (j-l).
    Variables have been normalized to mean values for better comparison amongst the runs.
    $R_g$ and $R_f$ show the Spearman rank correlation values for the subglacial and the subaerial outputs, respectively.
    Plots are shown with data and outputs at $15$\unit{min} intervals.
  }
  \label{fig:Qw_vari}
\end{figure}

Each water discharge value in the subaerial channel results in a unique water velocity, shear stress, and width-integrated shear stress, creating a perfect rank correlation between the variables (Figure~\ref{fig:Qw_vari} a--f).
In the subglacial channel, however,  the outputs can vary substantially for a given water discharge and width-integrated shear stress varies with respect to the other two  outputs as well (Figure~\ref{fig:Qw_vari} a--f).
For instance, very high  water velocity values and shear stresses can occur across minimal water discharge at $2.5$\,\unit{m}$^3$\,\unit{s}$^{-1}$ to the maximum water discharge at over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
In \icesheet, mean values of water velocity can occur at water discharges between roughly $150$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ and $310$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.

When considering the subglacial channel's evolving width, width-integrated shear stress  across the channel generally increases with water discharge, with improved rank correlation compared to water velocity or shear stress (Figure~\ref{fig:Qw_vari} a--f).
Yet even the width-integrated shear stress  can vary substantially, with the highest values occurring at water discharge values ranging from roughly $11$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ to over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
The variability in width-integrated shear stress is less pronounced in the \icesheet case, with reduced variability in water discharge (Figure~\ref{fig:Qw_vari} c, f).
The persistence of high water velocity, shear stress, and width-integrated shear stress suggests that substantial sediment transport may occur across much of the water discharge range in the subglacial system (Figure~\ref{fig:Qw_vari} a--f).

\begin{figure}[h]
  \centering
    \includegraphics[width=0.9\linewidth]{Fig4.png}
    \caption{Variability (measured by the standard deviation) in water velocity, shear stress, and width integrated shear stress for different time aggregations across the range of subglacial and subaerial channel shapes and friction factors.
      Shaded areas denote the minimum and maximum of the $100$ parameter combinations tested for each of the systems (Section~\ref{sect:imp}).
      Solid lines denote  mean values.
      Top row shows results from the \alpine{} case, while bottom row show outputs from the \icesheet{} case
      Time intervals shown with markers denote $15$\,\unit{min}, $1$\,\unit{hr}, $12$\,\unit{hr}, $1$\,\unit{d}, $5$\,\unit{d}, $10$\,\unit{d}, and $15$\,\unit{d} aggregation periods.
      Text in columns shows rank correlation between the variable and sediment discharge for the corresponding time period for the subglacial (right column) and subaerial (left column) channels.
    }
    \label{fig:multi_run}
  \end{figure}

The second application examines the impact of a variety of channel shapes and friction factors on the relationship between water discharge and sediment transport capacity in both systems.
Across the range of parameter values examined, variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one.
n \alpine{}, the variability in width-integrated shear stress is comparable between the two systems, due to the variable water discharge (Figure~\ref{fig:multi_run}).
 However, in \icesheet{}, with lower water discharge variability, the variability in width-integrated shear stress remains higher in the subglacial system across the time aggregation periods.

It is prescient to evaluate the time scales over which the subglacial channels may respond to changing water discharge to understand the influence of variability in sediment transport capacity.
In both cases, the variability in velocity, shear stress, and width-integrated shear stress decrease after $1-5$ day aggregations (Figure~\ref{fig:multi_run}).
Yet, substantial differences in variability persist in periods of up to $15$ days, as the variability in aggregations periods longer than this might be conflated with seasonal variations in water discharge.
Note that lags and time of peak sediment transport conditions still occur over these periods (Figure~\ref{fig:model_outs}), while the variability in sediment transport parameters decreases with larger time aggregation (Figure~\ref{fig:multi_run}).
\ian{add sentence here}

@@

When comparing the outputs with suspended sediment discharge measurements, the subaerial parameterizations generally show a better correlation with sediment discharge compared to subglacial outputs across the tested parameters (Figures~\ref{fig:Qw_vari} g--l and \ref{fig:multi_run}).
The rank correlation amongst the  subaerial parameterization's three output variables remains fairly constant despite the variable impact of channel shape and consistent scaling with water discharge (Figure \ref{fig:Qw_vari} and Section~\ref{sect:fluv}).
In subglacial channels, velocity and shear stress perform substantially worse in terms of rank correlation with sediment discharge compared to width-integrated shear stress (Figures~ \ref{fig:multi_run}).
In both \alpine{} and \icesheet{}, the  subglacial rank correlation approaches that of subaerial channels over longer time aggregation periods (Figures~ \ref{fig:multi_run} and \ref{fig:model_outs_1hr} to \ref{fig:model_outs_15day}).

\section{Discussion}
\subsection{Sediment capacity differences between the channel types}
\label{sect:dis_qsc}
Quantities that control sediment transport capacity, such as water velocity, shear stress, and width-integrated shear stress, vary differently in response to water discharge in subglacial channels compared to subaerial ones (Figure~\ref{fig:Qw_vari}).
In sediment transport relationships that evaluate sediment discharge capacity, such as in \citeA{meyer1948} or \citeA{engelund1967}, shear stress is scaled to the power of $\frac{3}{2}$ or $\frac{5}{2}$, respectively.
The exponent greater than $1$ magnifies sediment discharge variability beyond the variable sediment transport parameters described above (Figure~\ref{fig:multi_run}).
Note that sediment transport relationships describe sediment flux per unit width, so that the integrated capacity across channel occurs after application of the exponent, unlike the width-integrated shear stress evaluation here \cite<Equations\,\ref{eq:tautg} and \ref{eq:tautf}; c.f.>{tucker1997}.
Note also that both subglacial and subaerial parameterizations here assume that the distribution of velocity and shear stress is homogeneous across the channel bed, despite the limitations of this assumption \cite<e.g.>{yager2018}.

>>>>

%% To submit your paper:
\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage{placeins}
\usepackage{amsmath}
\usepackage{makecell}

@@

\title{Natural variations in water discharge quantity has an inconsistent impact on sediment transport capacity in subglacial channels}

\authors{Ian Delaney\affil{1},  Andrew J. Tedstone\affil{2},   Mauro A. Werder\affil{3,4},  Daniel Farinotti\affil{3,4} }

@@

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days to months, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated with increased or decreased water velocity $v$ \cite{alley1997} because the size of the channel $S$ changes on a much longer timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
%It follows that an increase in water discharge $Q$ will cause the shear stress across the channel $\tau_t$ to respond with water velocity $v$, as opposed to subaerial channels where channel width $w_c$ evolves as well.

  \begin{figure}[h]
  \centering
    \includegraphics[width=0.8\linewidth]{Fig1.pdf}
    \caption{Sketch for the different responses that subglacial and subaerial channels to increased water discharge.
      Water velocity magnitudes in the subglacial, $v_g$, and subaerial, $v_f$, channels are shown by arrow length.
      $S_g$ and $S_f$ represent the wetted area in subglacial and subaerial channels, respectively.
      The subglacial channel widths $w_{cg}$ remain unchanged, while the subaerial channel widths $w_{cf}$ evolve. Shear stress $\tau$ is responsible for the mobilization of sediment.
      %Note that in the subaerial channel parameterization, channels have rectangular shapes with their width much larger than their depth (Section~\ref{sect:fluv}).
    }
    \label{fig:cartoon}
  \end{figure}
% \ian{Come back to Daniel's comments}

As a result, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences are implicitly included in a wide range of available models that  quantify sediment transport in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>{perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channels \cite{alley1997}.
%Furthermore, thorough observations of water velocity and channel width in subglacial channels are limited and, to these authors' knowledge, not contrasted with subaerial conditions at the glacier margins. %\mauro{self-cite here?}

@@

The Leverett Glacier in Greenland (\icesheet) serves as the ice sheet setting.
Water discharge and suspended sediment concentration data were collected roughly $2$\unit{km} downstream from the terminus between June and August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used from \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.
Two gaps (14.07.2012 to 15.07.2012 and 16.07.2012 to 17.07.2021) in suspended sediment data interpolated using a linear spline.

@@

With wetted area $S_g$, the shear stress, $\tau_g$, between the water and the channel bed is determined through the Darcy-Weisbach formulation
\begin{equation}
  \label{eq:tau}
  \tau_g=\frac{1}{8}\,f_r\,\rho_w\,v_g^2,
\end{equation}
%
where $v_g = \frac{Q}{S_g}$ is the water velocity.

@@

where  channel floor width $w_{cg}$ is
\begin{equation}
  \label{eq:dh2wc}
  w_{cg} = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S_g}{\beta -\sin \beta}}.
\end{equation}

@@

To parameterize the shear stress of water flowing across sediments in the subaerial channel, the hydraulics parameterization presented in \citeA{tucker1997} is implemented.
Using mass conservation and the Darcy-Weisbach relationship as well as assuming that the channel is wide so that the hydraulic radius is well approximated by the flow depth, then
the shear stress $\tau_f$ at the river bed is
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q}{w_{cf}} \Big)^{\frac{2}{3}} \,\nabla z_c^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\nabla z_c$ is the channel slope, and $f_p$ is the friction factor for subaerial channels.
Channel width $w_{cf}$ is
\begin{equation}
  \label{eq:wcf}
  w_{cf} = k \, Q^{\alpha},
\end{equation}
%
where $k$ is a constant and $\alpha=\frac{1}{2}$ is a commonly chosen exponent \cite{leopold1953}.
Water velocity, $v_f$, is given by rearranging Equation \ref{eq:tau} as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_p\,\rho_w}}.
\end{equation}
%
As above, the width-integrated shear stress is
\begin{equation}
    \label{eq:tautf}
    \tau_{tf}=w_{cf}\,\tau_f.
  \end{equation}

@@

The parameterizations above are applied to hydrological records from the Fieschergletscher (\alpine) and the Leverett Glacier (\icesheet).
The subglacial and subaerial parameterizations represent the subglacial flow of water exiting a glacier and transitioning to subaerial flow.
Outputs of the parameterizations are meant to represent generalizable sediment transport characteristics from these hydrographs, rather than actual hydraulic conditions.
To generalize these cases, \alpine{}  is exemplified by relatively thin ice thickness ($h_{ice}$: $225$\,\unit{m}), low water discharge ($\sim\,10$\,\unit{m}$^3$\,\unit{s}$^{-1}$) and high diurnal variability in water discharge (Figure~\ref{fig:model_outs}, a).
\icesheet{}  is exemplified by thick ice  ($h_{ice}$: $700$\,\unit{m}), high water discharge ($\sim\,300$\,\unit{m}$^3$\,\unit{s}$^{-1}$)  and low diurnal variability in water discharge (Figure~\ref{fig:model_outs}, e).

For each glacier, the parameterizations are first applied to a reference test case for each glacier that assumes  subglacial channel with $\beta=\frac{\pi}{2}$.
Friction factors $f_r$ and $f_p$ are tuned so that the parameterizations reproduce reasonable water velocities for both \alpine and \icesheet{}\cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
In both subaerial channels, the bed slope is $\nabla z_c = 0.05$.
% Water velocity, shear stress, and width-integrated shear stress in the subglacial and subaerial channels from this formulation are presented below.

Second, to further characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels, $100$  parameterizations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$).
Runs with parameter combinations are culled if their mean subglacial water velocity over the season lies outside of the range between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies outside of the range $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
Variability of an output is represented by the  standard deviation of an output over the course of the season.

Lastly, outputs are compared with the sediment discharge records of the glaciers using Spearman rank correlation (i.e. correlation depends on the ordering of values from the outputs and the observations).
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.
Outputs and observations at $15$\,\unit{min} intervals are aggregated across timescales of up to $15$\,\unit{days} by averaging values one half of the timescale before and after a given point time.

@@

\subsection{Role of water discharge in sediment transport capacity}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig2.pdf}
  \caption{Parameterization outputs resulting from the hydrographs shown in panels a and e for \alpine cases (a-d) and \icesheet{} (e-h).
    Gray (black) lines in a and e represent the subglacial channel size (water discharge).
    In other panels, blue lines represent outputs from the subglacial channel, while orange lines represent the subaerial channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows denote some events where  variables peak in the subglacial channel prior to the subaerial one.
    Insets in e-h show the  peak melt event \icesheet{} denoted by the shaded area.
    Note that the x and y axes are different for \alpine{} and \icesheet{}.
      }
  \label{fig:model_outs}
\end{figure}

Following the parameterizations defined in Sections~\ref{sect:sub_mode}~and~\ref{sect:fluv}, the water velocity, shear stress, and width-integrated shear stress exhibit different seasonal evolutions and peaks.
Water velocity and shear stress represent the potential for sediment mobilization, and width-integrated shear stress represents a proxy for the total sediment transport capacity across the channel bed (Figure~\ref{fig:model_outs}).

Over the course of the season, large peaks in water velocity and shear stress, along with width-integrated shear stress, occur in response to water discharge variations in both \alpine{} and \icesheet{}.
In the subglacial channel, peaks generally occur over the period when water discharge increases at the fastest rate (Figure~\ref{fig:model_outs}).
Alternatively, in the subaerial channel, peaks in these sediment transport variables occur when water discharge is highest.
The timing of the greatest rate of change in water discharge occurs before the peaks in water discharge.
Furthermore, the subglacial outputs are generally larger than the subaerial values, thus more efficient at transporting sediment.
Here, sediment could be deposited in the proglacial area as the water depressurizes, then remobilized when sediment transport capacity peaks in the subaerial system.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig3.png}
  \caption{
    Relationship between water discharge and velocity, shear stress, and width-integrated shear stress for \alpine{} (a-c) and \icesheet (d-f).
    These outputs' relationship with measured sediment discharge is given for \alpine in (g-i) and \icesheet (j-l).
    Variables have been normalized to mean values.
    $R_g$ and $R_f$ show the Spearman rank correlations for the subglacial and the subaerial outputs, respectively.
    Plots are shown with data and outputs at $15$\unit{min} intervals.
  }
  \label{fig:Qw_vari}
\end{figure}

Each water discharge value in the subaerial channel results in a unique water velocity, shear stress, and width-integrated shear stress, creating a perfect rank correlation between the variables (Figure~\ref{fig:Qw_vari} a--f).
In the subglacial channel, however, the persistence of high water velocity, shear stress, and width-integrated shear stress across a range of discharges suggests that substantial sediment transport may occur across much of the season (Figure~\ref{fig:Qw_vari} a--f).
For instance, very high  water velocity values and shear stresses can occur across minimal water discharge at $2.5$\,\unit{m}$^3$\,\unit{s}$^{-1}$ to the maximum water discharge at over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
In \icesheet, mean values of water velocity can occur at water discharges between roughly $150$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ and $310$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
When considering the subglacial channel's evolving width, width-integrated shear stress  across the channel generally increases with water discharge, with improved rank correlation compared to water velocity or shear stress (Figure~\ref{fig:Qw_vari} a--f).
Yet even the width-integrated shear stress  can vary substantially, with the highest values occurring at water discharge values ranging from roughly $11$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ to over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
The variability in width-integrated shear stress is less pronounced in the \icesheet case, with reduced variability in water discharge (Figure~\ref{fig:Qw_vari} c, f).

\begin{figure}[h]
  \centering
    \includegraphics[width=0.9\linewidth]{Fig4.png}
    \caption{Variability (standard deviation) in water velocity, shear stress, and width integrated shear stress for different time aggregations across the range of subglacial and subaerial channel shapes and friction factors.
      Shaded areas denote the minimum and maximum of the $100$ parameter combinations tested (Section~\ref{sect:imp}).
      Solid lines denote  mean values.
      %Top row shows results from the \alpine{} case, while bottom row show outputs from the \icesheet{} case
      Time intervals shown with markers denote $15$\,\unit{min}, $1$\,\unit{hr}, $12$\,\unit{hr}, $1$\,\unit{d}, $5$\,\unit{d}, $10$\,\unit{d}, and $15$\,\unit{d} aggregation periods.
      Text in columns shows rank correlation between the variable and sediment discharge for the corresponding time period for the subglacial (right column) and subaerial (left column) channels.
    }
    \label{fig:multi_run}
  \end{figure}

The second application examines the impact of a variety of channel shapes and friction factors on the relationship between water discharge and sediment transport capacity.
Across the range of parameter values examined, variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one.
In \alpine{}, the variability in width-integrated shear stress is comparable between the two systems, due to the variable water discharge (Figure~\ref{fig:multi_run}).
However, in \icesheet{}, with lower water discharge fluctuations, the variability in width-integrated shear stress remains higher in the subglacial system across the time aggregation periods.

It is prescient to evaluate the time scales over which the subglacial channels may respond to changing water discharge to understand the influence of variability in sediment transport capacity.
In both cases, the variability in velocity, shear stress, and width-integrated shear stress decrease after $1-5$ day aggregations (Figure~\ref{fig:multi_run}).
Yet, substantial differences in variability persist in periods of up to $15$ days, as the variability in aggregations periods longer than this might be conflated with seasonal variations in water discharge.
Note that lags and time of peak sediment transport conditions still occur over these periods (Figure~\ref{fig:model_outs}), while the variability in sediment transport parameters decreases with larger time aggregation (Figure~\ref{fig:multi_run}).

@@

When comparing the outputs with suspended sediment discharge measurements, the subaerial parameterizations generally correlate better with sediment discharge compared to subglacial outputs across the tested parameters (Figures~\ref{fig:Qw_vari} g--l and \ref{fig:multi_run}).
The rank correlation amongst the subaerial parameterization's outputs remains fairly constant despite the variable impact of channel shape and consistent scaling with water discharge (Figure \ref{fig:Qw_vari} and Section~\ref{sect:fluv}).
In subglacial channels, velocity and shear stress perform substantially worse in terms of rank correlation with sediment discharge compared to width-integrated shear stress (Figures~ \ref{fig:multi_run}).
In both \alpine{} and \icesheet{}, the subglacial rank correlation approaches that of subaerial channels over longer time aggregation periods (Figures~ \ref{fig:multi_run} and \ref{fig:model_outs_1hr} to \ref{fig:model_outs_15day}).

\section{Discussion}
\subsection{Sediment capacity differences between the channel types}
\label{sect:dis_qsc}

Quantities that control sediment transport capacity, such as water velocity, shear stress, and width-integrated shear stress, vary differently in response to water discharge in subglacial channels compared to subaerial ones (Figure~\ref{fig:Qw_vari}).
In sediment transport relationships that evaluate sediment discharge capacity, such as in \citeA{meyer1948} or \citeA{engelund1967}, shear stress is scaled to the power of $\frac{3}{2}$ or $\frac{5}{2}$, respectively.
The exponent greater than $1$ magnifies sediment discharge variability beyond the variable sediment transport parameters described above (Figure~\ref{fig:multi_run}; Table~\ref{tab:Qs}).
Note that sediment transport relationships describe sediment flux per unit width, so that the integrated capacity across channel occurs after application of the exponent, unlike the width-integrated shear stress evaluation here \cite<Equations\,\ref{eq:tautg} and \ref{eq:tautf}; c.f.>{tucker1997}.
%Note also that both subglacial and subaerial parameterizations here assume that the distribution of velocity and shear stress is homogeneous across the channel bed, despite the limitations of this assumption \cite<e.g.>{yager2018}.

% The rapid increases and decreases in subglacial sediment transport capacity compounds the variability in sediment transport already present in subaerial systems \cite<Figure~\ref{fig:Qw_vari}; >{williams1989,jerolmack2010}.
Sediment transport capacity in subglacial channels rapidly responds to changes in water discharge compared to their subaerial counter parts \cite{alley1997}.
In subglacial R\"othlisberger channels where water discharge evolves more slowly than channel size or subaerial channels the width-integrated shear stress response to water discharge is $ \tau_{tg}\, \propto\,  Q^{\sim\frac{4}{5}}$ (Table\,\ref{tab:eqs1}).
However, subglacial channels rarely operate in a steady state \cite<e.g.>{gimbert2016}.
For sufficiently short time periods, channel width and wetted area are fixed in the subglacial system (assuming pipe flow) so that the response of width-integrated shear stress is $ \tau_{tg}\, \propto\,  Q^{2}$ (Table\,\ref{tab:eqs1}).
Appendix~\ref{sect:scaling} provides a complete description of sediment transport capacity in different drainage conditions.

The strong correlation between hydraulics and water discharge in \icesheet{} could result from being closer to an equilibrium between variations in the channels' wetter area and water discharge with smaller diurnal variations in amplitude (Figures~\ref{fig:model_outs} and ~\ref{fig:Qw_vari}).
Conversely, the correlation between hydraulics and water discharge in \alpine is smaller due to the larger discharge variations. 
In both cases, hydraulics could trend toward equilibrium by averaging the discharge over $1-5$ days, when the variability in subglacial channels approaches that of subaerial one (Figure~\ref{fig:multi_run}).

In subglacial channels, peaks in water velocity, shear stress and width-integrated shear stress occur with the fastest rate of water discharge increase relative to the channel's growth (Figure~\ref{fig:model_outs}).
In turn, subglacial channels grow with increasing water discharge, effectively reduces sediment transport capacity with the stabilization of water discharge at its peak.
Similar feedback occurs in subaerial systems \cite{phillips2016}, albeit at timescales longer than days \cite{gimbert2016}.
